---
title: "Gene Body Analysis"
output:
  html_document:
    df_print: paged
    toc: yes
  html_notebook:
    code_folding: hide
    toc: yes
---


```{r}
source("/gpfs/commons/groups/lappalainen_lab/jeinson/myPackages.R")
```

*Motivation:* If we look at variants in the gene bodies of all of the relevant genes, in addition to variants +/- 10kb of the TSS, can we get a better enrichment of rare variants?

```{r}
gene_attrs <- read_tsv("data/p05_relevant_genes_start_end.txt")
gene_attrs <- gene_attrs[!duplicated(gene_attrs$`Gene stable ID`),]
write_tsv(gene_attrs, "data/p05_relevant_genes_unique.txt")
```

What does the distribution of gene lengths look like?
```{r}
lengths <- gene_attrs$`Gene end (bp)` - gene_attrs$`Gene start (bp)`
hist(log10(lengths), main = "Historgram of log gene lengths")
abline(v = 4, col = "red", cex = 3)
summary(lengths)
```


How far is a transcription start site from the gene body start site, on average?
```{r}
TSS_dist <- gene_attrs$`Transcription start site (TSS)` - gene_attrs$`Gene start (bp)`
hist(log10(TSS_dist), main = "Histogram of log distance from TSS to gene start")
abline(v = 4, col = "red", cex = 3)

```

# .5% MAF or less SNVs

## Perform annotation processing workflow
```{r}
alt_allele_mismatch_correction <- 
  function(mat){
    
    mat[mat == -1] <- 0
    
    apply(mat, 1, function(r){
      if (sum(r == 2) > length(r) / 200) {# Maintain a 0.005 MaF
        r[r == 2] <- 0
        return(r)
      }
      else return(r)
    }) %>% t 
  }
```

```{r}
setwd("/gpfs/commons/groups/lappalainen_lab/jeinson/projects/aneva_river/gene_body_analysis/MAF005_gb_SNVs/")
all_genes <- list.files() %>% str_extract("ENSG\\d+") %>% unique

MAF_005_files <- list()

for(gene in all_genes){
  indv <- read_lines(paste0(gene, ".012.indv"))
  pos <- read_lines(paste0(gene, ".012.pos")) %>% str_replace_all("\t", ":")
  MAF_005_files[[gene]] <- stfu(read_delim(paste0(gene, ".012.t"), delim = " ", col_names = indv, skip = 1, progress = FALSE))
  stfu(rownames(MAF_005_files[[gene]]) <- pos)
}


MAF_005_files <- MAF_005_files %>% map(alt_allele_mismatch_correction)
```

```{r}
observed_snps <- do.call("rbind", MAF_005_files) %>% 
  rowSums %>% 
  is_greater_than(0) %>% which %>% names
```

```{r}
pVals <- read_tsv("/gpfs/commons/groups/lappalainen_lab/jeinson/projects/aneva_river/fourTiss_gb_analysis/fourTiss_AD_mean_pVals.tsv.gz")
pVals %<>% rename(GeneName = GENE_ID) %>% rename(AD_pVal = geo_means)
```

```{r}
rare_snv_enrichment <- 
  MAF_005_files %>%
  map(colSums) %>%
  map(as.data.frame) %>%
  map(~.x %>% rownames_to_column("SubjectID")) %>%
  bind_rows(.id = "GeneName") %>%
  rename(RareSNVs = `.x[[i]]`)

rare_snv_enrichment %<>% left_join(pVals %>% select(GeneName, SubjectID, AD_pVal)) %>% filter(!is.na(AD_pVal))
rare_snv_enrichment %<>% mutate(sig = as.numeric(AD_pVal < 0.05))
rare_snv_enrichment %<>% mutate(has_SNV = as.numeric(RareSNVs > 0))

to_keep <- rare_snv_enrichment %>% split(.$GeneName) %>% map("sig") %>% map_dbl(sum) %>% is_greater_than(0) %>% which %>% names

rare_snv_enrichment %<>% filter(GeneName %in% to_keep)

#Take a look at what we've got now
MAF005_enrichment <- rare_snv_enrichment
MAF005_enrichment
```

What does the distribution of AD pVals look like? We can decide on a cutoff from there. 
```{r}
hist(rare_snv_enrichment$AD_pVal, main = "Histogram of AD pValues across 800ish genes", xlab = "pVal")
hist(rare_snv_enrichment$AD_pVal[rare_snv_enrichment$AD_pVal < 0.05], main = "Historam of AD pValues less than 0.05", xlab = "pVal")
```

Make a Venn Diagram of significant genes and nearby rare variants
```{r}
library(limma)

a <- vennCounts(rare_snv_enrichment %>% select(sig, has_SNV))
vennDiagram(a, circle.col = c("yellow", "blue"))
```

$$ p(Sig) = 0.016 $$
$$ p(has\ SNV) = 0.398 $$
$$ p(has\ SNV | Sig) = 0.43 $$
$$ p(Sig | has\ SNV) = 0.018 $$
$$ p(Sig | no\ SNV ) = 0.0154$$

Enrichment:
$$ \frac{p(sig | has\ SNV)}{p(sig | no\ SNV)} = 1.166$$

# 1% MAF or less SNVs
```{r}
setwd("/gpfs/commons/groups/lappalainen_lab/jeinson/projects/aneva_river/gene_body_analysis/MAF01_gb_SNVs/")
all_genes <- list.files() %>% str_extract("ENSG\\d+") %>% unique

MAF_01_files <- list()

for(gene in all_genes){
  indv <- read_lines(paste0(gene, ".012.indv"))
  pos <- read_lines(paste0(gene, ".012.pos")) %>% str_replace_all("\t", ":")
  MAF_01_files[[gene]] <- stfu(read_delim(paste0(gene, ".012.t"), delim = " ", col_names = indv, skip = 1, progress = FALSE))
  stfu(rownames(MAF_01_files[[gene]]) <- pos)
}


MAF_01_files <- MAF_01_files %>% map(alt_allele_mismatch_correction)
```


```{r}
pVals <- read_tsv("/gpfs/commons/groups/lappalainen_lab/jeinson/projects/aneva_river/fourTiss_gb_analysis/fourTiss_AD_mean_pVals.tsv.gz")
pVals %<>% rename(GeneName = GENE_ID) %>% rename(AD_pVal = geo_means)
```

```{r}
rare_snv_enrichment <-
  MAF_01_files %>%
  map(colSums) %>%
  map(as.data.frame) %>%
  map(~.x %>% rownames_to_column("SubjectID")) %>%
  bind_rows(.id = "GeneName") %>%
  rename(RareSNVs = `.x[[i]]`)

rare_snv_enrichment %<>% left_join(pVals %>% select(GeneName, SubjectID, AD_pVal)) %>% filter(!is.na(AD_pVal))
rare_snv_enrichment %<>% mutate(sig = as.numeric(AD_pVal < 0.05))
rare_snv_enrichment %<>% mutate(has_SNV = as.numeric(RareSNVs > 0))

to_keep <- rare_snv_enrichment %>% split(.$GeneName) %>% map("sig") %>% map_dbl(sum) %>% is_greater_than(0) %>% which %>% names

rare_snv_enrichment %<>% filter(GeneName %in% to_keep)

#Take a look at what we've got now
MAF01_snv_enrichment <- rare_snv_enrichment
MAF01_snv_enrichment
```

Make a Venn Diagram
```{r}
library(limma)
a <- vennCounts(MAF01_snv_enrichment %>% select(sig, has_SNV))
vennDiagram(a, circle.col = c("yellow", "blue"))
```

$$ p(Sig) = 0.016 $$
$$ p(has\ SNV) = 0.517 $$
$$ p(has\ SNV | Sig) = 0.57 $$
$$ p(Sig | has\ SNV) = 0.018 $$
$$ p(Sig | no\ SNV) = 0.015 $$
Enrichment:

$$ \frac{p(Sig | has\ SNV)}{p(Sig | no\ SNV)} = 1.2 $$

